Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-13 linear_adia.tex
corrected generalized bcrit

Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
approximates the dispersion relation Eq. \ref{relax_disp} at marginal
stability with $\khat\gg 1$. We consider modes with $M=0$ (the
fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
1$ such that $|\hat{\omega}\khat|$ is $O(1)$ because $\khat\gg 1$. We
also assume $|q|$ and $\beta$ are $O(1)$ or less. For a thin disk
($\epsilon \ll 1$), we have approximately
\begin{align}
 0 = &M(M+1) + \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2\notag\\
   &-  \left(\hat{\omega}\khat\right)^4,\label{relax_cond_simp1}\\
   0= &M(M+1) + \frac{\beta}{\epsilon q} (1-\gamma)(1+2M)^2
   \left(\hat{\omega}\khat\right) + \left(\hat{\omega}\khat\right)^2. \label{relax_cond_simp2}
\end{align}
(We explain this in Appendix \ref{disp_neut_limit}.)

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and the critical thermal
timescale $\beta$ for marginal stability. If $\beta\ll 1$,
then a simple solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq M+1, 
\end{align}
which is the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Then  
\begin{align}\label{bcrit_gen}
  \beta = \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{(1+M)^{3/2}}{(1+2M)^2} \leq \beta_\mathrm{crit} 
\end{align}
where 
\begin{align}\label{iso_vsi_cond}
  \beta_\mathrm{crit}  \equiv \frac{\epsilon|q|}{\gamma-1}
\end{align}
is the critical thermal timescale for the fundamental mode
($M=0$). However, according to Eq. \ref{bcrit_gen}  the critical
thermal timescale decreases with increasing  $M$. We thus expect
higher order modes to be stabilized more rapidly by finite thermal
relaxation. When $\beta>\beta_\mathrm{crit}$, all modes
(satisfying the above assumptions) are stabilized. We can interpret
$\beta_\mathrm{crit}$ as the thermal timescale below which
perturbations are effectively isothermal to allow the VSI.  We will confirm
Eq. \ref{iso_vsi_cond} numerically in \S\ref{bcrit_num_test}.

(We remark that the critical thermal timescale $\beta_\mathrm{crit}$
for the fundamental mode may be obtained more readily with fewer
assumptions by setting $M=0$ in
Eq. \ref{relax_cond1}---\ref{relax_cond2} from the onset.)

>>>>

Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
approximates the dispersion relation Eq. \ref{relax_disp} at marginal
stability with $\khat\gg 1$. We consider modes with $M=0$ (the
fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
1$ such that $|\hat{\omega}\khat|$ is $O(1)$ because $\khat\gg 1$. We
also assume $|q|$ and $\beta$ are $O(1)$ or less. For a thin disk
($\epsilon \ll 1$), we have approximately
\begin{align}
 0 = &M(M+1)\left(1-\epsilon^2q^2\khat^2\right) + \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2\notag\\
   &-  \left(\hat{\omega}\khat\right)^4,\label{relax_cond_simp1}\\
   0= &2M(M+1) + \frac{\beta}{\epsilon q} (1-\gamma)(1+2M)^2
   \left(\hat{\omega}\khat\right) + \left(\hat{\omega}\khat\right)^2. \label{relax_cond_simp2}
\end{align}
(We explain this in Appendix \ref{disp_neut_limit}.)

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and the critical thermal
timescale $\beta$ for marginal stability. If $\beta\ll 1$ and the disk
is sufficiently thin so that $|\epsilon q \khat |\ll 1$
then a simple solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq M+1, 
\end{align}
which is the dispersion relation for low-frequency waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Then  
\begin{align}\label{bcrit_gen}
  \beta = \frac{\epsilon|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)} \leq \beta_\mathrm{crit} 
\end{align}
where 
\begin{align}\label{iso_vsi_cond}
  \beta_\mathrm{crit}  \equiv \frac{\epsilon|q|}{\gamma-1}
\end{align}
is the critical thermal timescale for the fundamental mode
($M=0$). However, according to Eq. \ref{bcrit_gen}  the critical
thermal timescale decreases with increasing  $M$. We thus expect
higher order modes to be stabilized more rapidly by finite thermal
relaxation. When $\beta>\beta_\mathrm{crit}$, all modes
(satisfying the above assumptions) are stabilized. We can interpret
$\beta_\mathrm{crit}$ as the thermal timescale below which
perturbations are effectively isothermal to allow the VSI.  We will confirm
Eq. \ref{iso_vsi_cond} numerically in \S\ref{bcrit_num_test}.

(We remark that the critical thermal timescale for the fundamental
mode, $\beta_\mathrm{crit}$, may be obtained more readily without 
assuming $|\epsilon q \khat|\ll 1$ by setting $M=0$ in
Eq. \ref{relax_cond1}---\ref{relax_cond2} from the onset.)

### 2015-03-13 numerical.tex
corrected generalized bcrit

\begin{figure}
  \includegraphics[scale=0.345,clip=true,trim=0cm 0cm 3.1cm
  0cm]{figures/result2d_vel}\includegraphics[scale=0.345,clip=true,trim=1.9cm 0cm 0cm
  0cm]{figures/result2d_mom} 
  \caption{Visualization of the fundamental VSI mode in
    Fig. \ref{lowfreq_eigenfunc}. The color scale corresponds to the
    real pressure perturbation $W$ scaled by its maximum value.
    Arrows in the left panel correspond to the meridional velocity
    field $(\delta v_x,\delta v_z)$, and arrows in the right panel
    correspond to to the meridional momentum $\rho(\delta v_x,\delta
    v_z)$. 
    \label{lowfreq_eigenfunc_2d}
  }
\end{figure}

@@

\begin{figure}
  \includegraphics[scale=0.345,clip=true,trim=0cm 0cm 3.1cm
  0cm]{figures/result2d_vel_cool}\includegraphics[scale=0.345,clip=true,trim=1.9cm 0cm 0cm
  0cm]{figures/result2d_mom_cool} 
  \caption{Same as  Fig. \ref{lowfreq_eigenfunc_2d} but with a thermal
    relaxation timescale $\beta=0.1$. 
    \label{lowfreq_eigenfunc_2d_cool}
  }
\end{figure}

>>>>

\begin{figure}
%  \includegraphics[scale=0.345,clip=true,trim=0cm 0cm 2.5cm
%  0cm]{figures/result2d_vel}\includegraphics[scale=0.345,clip=true,trim=1.9cm 0cm 0cm
%  0cm]{figures/result2d_mom} 
  \includegraphics[width=\linewidth]{figures/result2d_iso}
  \caption{Visualization of the fundamental VSI mode in
    Fig. \ref{lowfreq_eigenfunc}. The color scale corresponds to the
    real pressure perturbation $\delta P$ scaled by its maximum value.
    Arrows correspond to the vector field 
    field $\sqrt{\rho}(\delta v_x,\delta v_z)$. Note that the horizontal scale has been scretched for clarity. 
    \label{lowfreq_eigenfunc_2d}
  }
\end{figure}

@@

\begin{figure}
  % \includegraphics[scale=0.345,clip=true,trim=0cm 0cm 2.5cm
  % 0cm]{figures/result2d_vel_cool}\includegraphics[scale=0.345,clip=true,trim=1.9cm 0cm 0cm
  % 0cm]{figures/result2d_mom_cool} 
  \includegraphics[width=\linewidth]{figures/result2d_cool}
  \caption{Same as  Fig. \ref{lowfreq_eigenfunc_2d} but with a thermal
    relaxation timescale $\beta=0.1$. 
    \label{lowfreq_eigenfunc_2d_cool}
  }
\end{figure}
